---
title: "CHASM Example Full Pipeline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CHASM Example Full Pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 12,
  fig.height = 8,
  eval = FALSE
)

find_package_root <- function(path) {
  current <- normalizePath(path, winslash = "/", mustWork = FALSE)

  repeat {
    description_path <- file.path(current, "DESCRIPTION")
    if (file.exists(description_path)) {
      return(current)
    }

    parent <- dirname(current)
    if (identical(parent, current)) {
      return(NULL)
    }
    current <- parent
  }
}

package_root <- find_package_root(getwd())
if (is.null(package_root)) {
  package_root <- find_package_root(dirname(knitr::current_input()))
}

if (!is.null(package_root) && requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(package_root, export_all = FALSE, quiet = TRUE)
} else if (requireNamespace("CHASM", quietly = TRUE)) {
  suppressPackageStartupMessages(library(CHASM))
} else {
  stop("CHASM is not installed and pkgload is unavailable.")
}

resolve_vignette_example <- function(file_name) {
  if (!is.null(package_root)) {
    source_path <- file.path(package_root, "inst", "extdata", file_name)
    if (file.exists(source_path)) {
      return(source_path)
    }
  }

  example_data_path(file_name)
}
```

This tutorial walks through the bundled end-to-end CHASM example using the
packaged example data. It mirrors the logic in
`tutorial/run_chasm_example_full_pipeline.R`, and uses a cached heatmap input
object so the final heatmap can be rendered from executed plotting code during
vignette build without rerunning the full pipeline.

```{r packages}
if (!requireNamespace("dplyr", quietly = TRUE) ||
    !requireNamespace("tidyr", quietly = TRUE) ||
    !requireNamespace("stringr", quietly = TRUE) ||
    !requireNamespace("rsvd", quietly = TRUE) ||
    !requireNamespace("pheatmap", quietly = TRUE)) {
  stop(
    "This vignette requires dplyr, tidyr, stringr, rsvd, and pheatmap."
  )
}
```

## Load bundled example data

```{r input-data}
sanitize_cell_ids <- function(x) {
  getFromNamespace("normalize_dnacopy_ids", "CHASM")(x)
}

chromosomes <- paste0("chr", c(1:22, "X", "Y"))

read_depth_file <- "Read_depth_matrix_chr1_50mb_2_5pct_Rep1.rds"
read_depth_chrom_file <- sub("\\.rds$", "_chrom.rds", read_depth_file)
spikein_cells_file <- "read_depth_spikein_cells_2mb_ALL.rds"

sample_name <- sub("^Read_depth_matrix_", "", sub("\\.rds$", "", read_depth_file))

read.depth <- readRDS(example_data_path(read_depth_file))
read.depth.chrom <- readRDS(example_data_path(read_depth_chrom_file))
cells.spikein <- readRDS(example_data_path(spikein_cells_file))
```

## Prepare the binned read-depth matrix

```{r prepare-wavelet-input}
rownames(read.depth) <- sanitize_cell_ids(rownames(read.depth))
read.depth$barcode <- rownames(read.depth)

bins <- setdiff(colnames(read.depth), "barcode")
chrom_bins <- bins[grepl("^chr", bins)]
ki_gl_bins <- bins[grepl("^(KI|GL)", bins)]
bins <- setdiff(chrom_bins, ki_gl_bins)
read.depth <- read.depth[, c(bins, "barcode")]
```

## Normalize and denoise with the wavelet workflow

```{r wavelet-workflow}
chrom.depth.per.cell <- normalize_depth(read.depth, bins)
wavelet.transform <- wavelet_transform(chrom.depth.per.cell, bins)
robust.pca <- robust_pca(wavelet.transform$mat.wavelet.transform)

residuals.mat <- inv_wavelet_transform(
  robust.pca$Sparse_Signal,
  wavelet.transform$chrom.informed.wavelet
)
svd.normalized.read.depth.mat <- inv_wavelet_transform(
  robust.pca$Expected_Normal,
  wavelet.transform$chrom.informed.wavelet
)

colnames(residuals.mat) <- rownames(wavelet.transform$mat.wavelet.transform)
colnames(svd.normalized.read.depth.mat) <- rownames(wavelet.transform$mat.wavelet.transform)
```

## Segment residual signal and assign bin-level copy-number states

```{r segment-and-assign}
segment.table <- segment_residuals(residuals.mat, alpha = 0.005)
cn_bin <- assign_cn_state(chrom.depth.per.cell, svd.normalized.read.depth.mat, segment.table)
```

## Run chromosome-level state assignment

```{r chromosome-level}
positions <- colnames(read.depth.chrom)
read.depth.chrom$ID <- sanitize_cell_ids(rownames(read.depth.chrom))
read.depth.chrom$celltype <- "unknown"
rownames(read.depth.chrom) <- read.depth.chrom$ID

cn_state.nb <- assign_cn_state.chrom(read.depth.chrom, positions)
```

## Merge segment-level and chromosome-level calls

```{r merge-calls}
cn_df.comb <- merge_calls(cn_bin, cn_state.nb)
cn_df.comb$ID_clean <- sanitize_cell_ids(cn_df.comb$ID)
cn_df.comb$chrom_name <- stringr::str_split_fixed(cn_df.comb$chrom_bin, "_", 3)[, 1]
```

## Prepare the heatmap matrix

```{r heatmap-matrix, eval = TRUE}
heatmap_input <- readRDS(resolve_vignette_example("chasm_vignette_heatmap_input.rds"))
cn_bin.wide <- heatmap_input$cn_bin.wide
annotation_row <- heatmap_input$annotation_row
bin_df_break <- heatmap_input$bin_df_break
```

## Plot the example heatmap

```{r heatmap, echo = TRUE, eval = TRUE, fig.width=5, fig.height=5}
pheatmap::pheatmap(
  cn_bin.wide,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  show_rownames = FALSE,
  show_colnames = FALSE,
  annotation_row = annotation_row,
  annotation_colors = list(Spike_in = c(Yes = "red", No = "blue")),
  gaps_col = bin_df_break$max_order,
  gaps_row = sum(annotation_row$Spike_in == "Yes")
)
```

## Save outputs if needed

The original script writes intermediate and final objects to disk. For an
interactive tutorial page, it is usually cleaner to leave the outputs in the
session, but the same objects are available if you want to save them:

```{r save-outputs, eval = FALSE}
wavelet_output <- list(
  cn_bin = cn_bin,
  segment.table = segment.table,
  chrom.depth.per.cell = chrom.depth.per.cell,
  residuals.mat = residuals.mat
)

# saveRDS(wavelet_output, paste0(sample_name, "_output_wavelet.rds"))
# saveRDS(cn_df.comb, paste0(sample_name, "_output_cn_df.comb.rds"))
# write.csv(cn_df.comb, paste0(sample_name, "_example_cn_calls.csv"), row.names = FALSE)
```
